% !TEX root = ../main.tex

% 专硕请取消下面注释
\makeatletter
  \def\hitsz@csubjecttitle{类别}
\makeatother

\hitszsetup{
  %******************************
  % 注意：
  %   1. 配置里面不要出现空行
  %   2. 不需要的配置信息可以删除
  %******************************
  %
  %=====
  % 秘级
  %=====
  statesecrets={公开},
  natclassifiedindex={TP399},
  intclassifiedindex={004.8},
  %
  %=========
  % 中文信息
  %=========
  ctitleone={基于多模态预训练},%本科生封面使用
  ctitletwo={蛋白质功能预测研究},%本科生封面使用
  ctitlecover={基于多模态预训练的蛋白质功能预测研究},%放在封面中使用，自由断行
  ctitle={基于多模态预训练的蛋白质功能预测研究},%放在原创性声明中使用
  %csubtitle={一条副标题}, %一般情况没有，可以注释掉
  cxueke={专业},
  cpostgraduatetype={专业},
  csubject={计算机技术},
  caffil={哈尔滨工业大学（深圳）},
  cauthor={崔博越},
  csupervisor={李君一教授},
  % 日期自动使用当前时间，若需指定按如下方式修改：
  cdate={2026年5月},
  % 指定第二页封面的日期，即答辩日期
  cdatesecond={2026年05月},
  cstudentid={23S151070},
  cstudenttype={同等学力人员}, %非全日制教育申请学位者
  %
  %=========
  % 英文信息
  %=========
  etitle={Research on Protein Function Prediction Based on Multimodal Pretraining},
  exueke={Engineering},
  esubject={Computer Technology},
  eaffil={Harbin Institute of Technology, Shenzhen},
  eauthor={Boyue Cui},
  esupervisor={Prof. Junyi Li},
  % 日期自动生成，若需指定按如下方式修改：
  edate={May, 2026},
  estudenttype={Master of Engineering},
  %
  % 关键词用"英文逗号"分割
  ckeywords={蛋白质功能预测, 零样本学习, 蛋白质语言模型, 多模态预训练},
  ekeywords={protein function prediction, zero-shot learning, protein language model, multimodal pretraining},
}

% 中文摘要
\begin{cabstract}

蛋白质功能预测是生物信息学领域的核心任务之一，旨在揭示蛋白质序列与生物学功能之间的映射关系。随着高通量测序技术的快速发展，蛋白质序列数据呈指数级增长，而经实验验证的功能注释仍十分有限。在此背景下，依靠计算方法开展自动化功能预测，已成为缓解注释不足的重要途径。现有深度学习方法通常将基因本体（Gene Ontology, GO）标签视为固定的分类标识，主要依赖序列模式进行封闭标签空间下的分类，较少利用蛋白质域描述和GO定义中蕴含的语义信息，在未见标签和长尾类别上泛化能力不足。本论文围绕"多模态语义建模"与"预训练增强"两条递进路径，分别从开放标签预测和端到端部署两个层面展开研究。

针对现有方法语义利用不充分、难以预测未见功能标签的问题，本论文提出了一种多模态零样本蛋白质功能注释方法MZSGO。该方法将蛋白质功能预测表述为蛋白质与GO术语之间的语义匹配问题，利用蛋白质语言模型ESM2提取序列进化特征，并借助文本嵌入模型对蛋白质域描述与GO功能定义进行统一语义编码。在此基础上，引入非对称特征丢弃策略增强模型对模态缺失场景的鲁棒性，设计自适应门控融合模块动态调节不同信息源的贡献权重，实现异构特征在共享潜在空间中的有效整合。实验结果表明，MZSGO在有监督指标上具有竞争力，并在零样本场景下显著优于现有方法，验证了多模态语义匹配范式的有效性。

为降低推理阶段对显式域语义输入的依赖，本论文进一步提出基于域感知适配器预训练的方法MZSGO-DA。该方法将轻量级适配器注入序列模型的Transformer层，通过"序列—域"和"序列—功能"双通道对比学习预训练，将域功能语义注入适配器参数中。下游阶段采用冻结与可训练适配器并行的双适配器策略，兼顾预训练先验保留与任务适配。实验结果表明，MZSGO-DA仅依赖序列输入便在多数指标上达到或超越MZSGO，尤其在分子功能本体上实现全面提升，显著降低了推理阶段的输入依赖。

本论文的研究表明，融合蛋白质语言模型、文本嵌入模型、域功能语义与基因本体标签定义，能够有效提升蛋白质功能预测在有监督和零样本场景下的准确性与泛化能力。从显式多模态语义融合到隐式域功能知识注入的递进式设计，为促进蛋白质功能自动化解析、降低对实验验证的依赖提供了新的思路与方法支持。

\end{cabstract}

% 英文摘要
\begin{eabstract}

Protein function prediction is a fundamental task in bioinformatics, aiming to uncover the mapping between protein sequences and biological functions. With the rapid growth of high-throughput sequencing, protein sequence data has expanded exponentially, whereas experimentally verified functional annotations remain scarce. As a result, the disparity between available sequences and verified functional annotations continues to grow, making computational prediction an essential solution for large-scale automatic annotation. Most existing deep learning approaches treat Gene Ontology (GO) terms as fixed class identifiers and mainly rely on sequence patterns for closed-set classification. Such a formulation underutilizes the rich semantic information contained in protein domain descriptions and GO definitions, which in turn limits generalization to unseen labels and long-tail categories. This dissertation addresses these challenges along two progressive directions: multimodal semantic modeling for open-label prediction and pretraining-based enhancement for end-to-end deployment.

To address the insufficient use of semantic knowledge and the inability to predict unseen functional labels, this dissertation proposes MZSGO, a multimodal zero-shot protein function annotation method. MZSGO formulates protein function prediction as a semantic matching problem between proteins and GO terms. It uses the protein language model ESM2 to extract evolutionary sequence representations and employs a text embedding model to encode protein domain descriptions and GO definitions into a unified semantic space. An asymmetric feature dropout strategy is introduced to improve robustness under missing-modality conditions, and an adaptive gated fusion module is designed to dynamically weight different information sources and integrate heterogeneous features in a shared latent space. Experimental results show that MZSGO achieves competitive performance on supervised benchmarks and substantially outperforms existing methods in zero-shot settings, validating the effectiveness of the multimodal semantic matching paradigm.

To reduce the reliance on explicit domain semantic input during inference, this dissertation further proposes MZSGO-DA, a method based on domain-aware adapter pretraining. MZSGO-DA injects lightweight adapters into the Transformer layers of the sequence encoder and performs dual-channel contrastive pretraining on sequence--domain and sequence--function pairs, thereby injecting domain-related functional semantics into the adapter parameters. In the downstream stage, a dual-adapter strategy is adopted, where one pretrained adapter is frozen to preserve prior knowledge and another trainable adapter is introduced for task-specific adaptation. Experimental results show that MZSGO-DA, requiring only sequence input at inference time, achieves comparable or superior performance to MZSGO on most metrics, with particularly notable improvements on the Molecular Function ontology, while substantially reducing the input dependency at inference time.

Overall, this dissertation demonstrates that integrating protein language models, text embedding models, domain-functional semantic information, and Gene Ontology label definitions can effectively improve the accuracy and generalization of protein function prediction under both supervised and zero-shot settings. The progressive design, from explicit multimodal semantic fusion to implicit domain-functional knowledge injection, provides new insights and methodological support for advancing large-scale automated protein function annotation and reducing the reliance on experimental validation.

\end{eabstract}
